A hazards approach to the biometric analysis of infant mortality

### 2023-03-16 paper2.Rmd
add cohort and period

@@

However, this data set is known to be of very high quality [@belu99], so we are inclined to 
think that the biometric model as it is described here simply is inappropriate during the eighteenth century.

>>>>

## The assumptions about exogenous mortality

*Exogenous* causes of death are due to social, cultural, economic, and environmental 
factors, and also accidents. An argument for the constant hazards and the uniform approaches to
exogenous infant mortality is that, after compensating for age via the log-cube transformation, 
these factors are random and evenly spread in time, like *white noise*. However,  this disregards 
seasonal variations in weather, and following that the spread of infectious diseases.

To illustrate this, data will be split in two ways, first by "birth month" (cohort data) and second by
"calendar month" (period data).

### Cohorts by birth season

```{r datain}
library(eha)
Sys.setlocale("LC_ALL", "en_US.utf8")
g <- function(x) (log(x+1))^3 * 365 / (log(366))^3
infa <- readRDS("Rdata/infant.rds")[, c("id", "sex", "birthdate", "enter", "exit", "event")]
infa$exit <- infa$exit / 365 ## NOTE!!
infa$bmonth <- months(infa$birthdate, abbreviate=TRUE)
infa$bmonth <- factor(infa$bmonth, levels = month.abb)
winter <- with(infa, bmonth %in% c("Dec", "Jan", "Feb"))
spring <- with(infa, bmonth %in% c("Mar", "Apr", "May"))
summer <- with(infa, bmonth %in% c("Jun", "Jul", "Aug"))
fall <- with(infa, bmonth %in% c("Sep", "Oct", "Nov"))
infa$bseason <- ""
infa$bseason[winter] <- "winter"
infa$bseason[summer] <- "summer"
infa$bseason[fall] <- "fall"
infa$bseason[spring] <- "spring"
infa$bseason <- factor(infa$bseason, levels = c("spring", "summer", "fall", "winter"))
infa$bdate <- toTime(infa$birthdate)
##infa$bdate <- (infa$bdate - floor(infa$bdate)) * 365
cuts <- seq(0, 1, by = 1 / 12)
comm <- data.frame(month = rep(month.abb, 100))
dat <- make.communal(infa, comm, start = 1861, period = 1 / 12,
                     surv = c("enter", "exit", "event", "bdate"))
dat$month <- factor(dat$month, levels = month.abb)
winter <- with(dat, month %in% c("Dec", "Jan", "Feb"))
spring <- with(dat, month %in% c("Mar", "Apr", "May"))
summer <- with(dat, month %in% c("Jun", "Jul", "Aug"))
fall <- with(dat, month %in% c("Sep", "Oct", "Nov"))
dat$season <- ""
dat$season[winter] <- "winter"
dat$season[spring] <- "spring"
dat$season[summer] <- "summer"
dat$season[fall] <- "fall"
dat$season <- factor(dat$season, levels = c("spring", "summer", "fall", "winter")) 
dat$enter <- 365 * dat$enter # NOTE!
dat$exit <- 365 * dat$exit # NOTE!
```

```{r plotcohort, fig.cap = "By birth month, cohorts.", fig.height = 9, eval = FALSE}
oldpar <- par(mfrow = c(4, 3))
for (mon in levels(dat$bmonth)){
    drawBB2(dat[dat$bmonth == mon, ], main = mon, printLegend = FALSE)
}
par(oldpar)
```

```{r plotcohort2, fig.cap = "By birth season, cohorts.", fig.height = 8}
oldpar <- par(mfrow = c(2, 2))
for (seas in levels(dat$bseason)){
    drawBB2(dat[dat$bseason == seas, ], main = seas, printLegend = FALSE)
}
par(oldpar)
```

We see that the post-neonatal part of the total cumulative hazards are more or
less convex (bend upward at the end) for those born in spring or summer, while 
the curves are concave for births in the fall or winter. Bad fits to "constant hazard"
and "uniform distribution".

### Periods by season

```{r plotperiod, fig.cap = "By calendar month, periods.", fig.height = 9, eval = FALSE}
oldpar <- par(mfrow = c(4, 3))
for (mon in levels(dat$bmonth)){
    drawBB2(dat[dat$month == mon, ], main = mon, printLegend = FALSE)
}
par(oldpar)
```

```{r plotperiod2, fig.cap = "By calendar season, periods.", fig.height = 8}
oldpar <- par(mfrow = c(2, 2))
for (seas in levels(dat$season)){
    drawBB2(dat[dat$season == seas, ], main = seas, printLegend = FALSE)
}
par(oldpar)
```
The periods are of course more homogeneous regarding weather and external threats,
and it shows off in that in all four periods the suggested models fit well, here shown by the
"constant hazards model" for exogenous mortality.

### Conclusion

The assumptions about constant hazard and uniform distribution of exogenous mortality 
are only applicable on population level, it can never hold for an individual. It can 
be thought of as an average over birth cohorts, where

@@

However, this data set is known to be of very high quality [@tbcl99], so we are inclined to 
think that the biometric model as it is described here simply is inappropriate during the eighteenth century.
